---
title: "Data Visualization in R with ggplot2"
author: "Mine Cetinkaya-Rundel"
date: "4/16/2019"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
```

# Data visualization

## Datasets

* Transit ride data
    + `daily`: daily summary of rides

* Durham registered voter data
    + `durham_voters`: one row per voter

```{r, message=FALSE, eval=TRUE} 
daily <- read_csv("data/daily.csv")
durham_voters <- read_csv("data/durham_voters.csv")
```

# Layer up!

```{r echo=FALSE}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(se = FALSE, method = "loess") +
  theme_minimal() +
  labs(x = "Ride date", y = "Number of rides", color = "Day of week",
       title = "Daily rides", subtitle = "Durham, NC") +
  scale_color_viridis_d()
```

## Your turn!

**Exercise:** Which of the four datasets does this visualization use? Determine 
which variable is mapped to which aesthetic (x-axis, y-axis, etc.) element 
of the dataset.

```{r echo=FALSE,}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(se = FALSE, method = "loess") +
  theme_minimal() +
  labs(x = "Ride date", y = "Number of rides", color = "Day of week",
       title = "Daily rides", subtitle = "Durham, NC") +
  scale_color_viridis_d()
```

## Step-by-step

```{r}
ggplot(data = daily)
```

```{r}
ggplot(data = daily, mapping = aes(x = ride_date, y = n_rides))
```

```{r}
ggplot(data = daily, mapping = aes(x = ride_date, y = n_rides)) + 
  geom_point()
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) + 
  geom_point()
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) + 
  geom_smooth()
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) + 
  geom_smooth(method = "loess")
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) + 
  geom_smooth(method = "loess", se = FALSE)
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) + 
  geom_smooth(method = "loess", se = FALSE) +
  scale_color_viridis_d()
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) + 
  geom_smooth(method = "loess", se = FALSE) +
  scale_color_viridis_d() +
  theme_minimal()
```

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(x = "Ride date", y = "Number of rides", color = "Day of week",
       title = "Daily rides", subtitle = "Durham, NC")
```

# Mapping

## Size by number of riders

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, size = n_riders)) +
  geom_point()
```

## Set alpha value

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, size = n_riders)) +
  geom_point(alpha = 0.5)
```

## Your turn!

**Exercise:** Using information from https://ggplot2.tidyverse.org/articles/ggplot2-specs.html add color, size, alpha, and shape aesthetics to your graph. Experiment. Do different things 
happen when you map aesthetics to  discrete and continuous variables?
What happens when you use more than one aesthetic?

```{r, eval=FALSE}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(x = "Ride date", y = "Number of rides", color = "Day of week",
       title = "Daily rides", subtitle = "Durham, NC")
```

## Mappings can be at the `geom` level

```{r}
ggplot(data = daily) +
  geom_point(mapping = aes(x = ride_date, y = n_rides))
```

## Different mappings for different `geom`s

```{r}
ggplot(data = daily, mapping = aes(x = ride_date, y = n_rides)) +
  geom_point() +
  geom_smooth(aes(color = day_of_week), method = "loess", se = FALSE)
```

## Set vs. map

- To **map** an aesthetic to a variable, place it inside `aes()`
```{r}
ggplot(data = daily, 
  mapping = aes(x = ride_date, 
                y = n_rides,
            color = day_of_week)) +
  geom_point() 
```

- To **set** an aesthetic to a value, place it outside `aes()`
```{r}
ggplot(data = daily, 
  mapping = aes(x = ride_date, 
                y = n_rides)) +
  geom_point(color = "red") 
```

# Syntax

## Data can be passed in

```{r}
daily %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
    geom_point()
```

## Parameters can be unnamed

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides)) +
  geom_point()
```

## Variable creation on the fly...

Color by weekday / weekend

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, 
       color = day_of_week %in% c("Sat", "Sun"))) +
  geom_point()
```


## Variable creation on the fly...

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, 
       color = day_of_week %in% c("Sat", "Sun"))) +
  geom_point() +
  labs(color = "Weekend")
```

## ... or passed in

```{r}
daily %>%
  mutate(day_type = if_else(day_of_week %in% c("Sat", "Sun"),
                            "Weekend",
                            "Weekday")) %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_type)) +
    geom_point()
```

# Common early pitfalls

## Mappings that aren't

```{r}
ggplot(data = daily) +
  geom_point(aes(x = ride_date, y = n_rides, color = "blue"))
```

## Mappings that aren't

```{r}
ggplot(data = daily) +
  geom_point(aes(x = ride_date, y = n_rides), color = "blue")
```

## Your turn!

**Exercise:** What is wrong with the following?

```{r eval = FALSE}
daily %>%
  mutate(day_type = if_else(day_of_week %in% c("Sat", "Sun"),
                            "Weekend",
                            "Weekday")) %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_type)) %>%
    geom_point()
```

## + and %>%

What is wrong with the following?

```{r error = TRUE}
daily %>%
  mutate(day_type = if_else(day_of_week %in% c("Sat", "Sun"),
                            "Weekend",
                            "Weekday")) %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_type)) %>%
    geom_point()
```

# Building up layer by layer

## Basic plot

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides)) +
  geom_point() 
```

## Two layers!

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides)) +
  geom_point() +
  geom_line()
```

## Iterate on layers

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides)) +
  geom_point() + 
  geom_smooth(span = 0.1) # try changing span
```

## The power of groups

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() + 
  geom_line()
```

## Now we've got it

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(span = 0.2, se = FALSE)
```

## Control data by layer

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = filter(daily,
                           !(day_of_week %in% c("Sat", "Sun")) 
                           & n_rides < 200),
             size = 5, color = "gray") +
  geom_point()
```

## Your turn!

**Exercise:** Work with your neighbor to sketch what the following plot will 
look like. No cheating! Do not run the code, just think through the code for 
the time being.

## Step by step

```{r}
low_weekdays <- daily %>%
  filter(!(day_of_week %in% c("Sat", "Sun")) & n_rides < 100)

low_weekdays
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point()
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  geom_point(data = low_weekdays, size = 5, color = "gray")
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = low_weekdays, size = 5, color = "gray") +
  geom_point()
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = low_weekdays, size = 5, color = "gray") +
  geom_point() +
  geom_text(data = low_weekdays, aes(y = n_rides, label = ride_date), 
            size = 2, color = "black")
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = low_weekdays, size = 5, color = "gray") +
  geom_point() +
  geom_text(data = low_weekdays, aes(y = n_rides + 15, label = ride_date), 
            size = 2, color = "black")
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = low_weekdays, size = 5, color = "gray") +
  geom_point() + 
  geom_text_repel(data = low_weekdays, 
                  aes(x = ride_date, y = n_rides, 
                      label = as.character(ride_date)), 
                  size = 3, color = "black")
```

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = low_weekdays, size = 5, color = "gray") +
  geom_point() + 
  geom_label_repel(data = low_weekdays, 
                  aes(x = ride_date, y = n_rides, 
                      label = as.character(ride_date)), 
                  size = 2, color = "black")
```

## Your turn!

**Exercise:** How would you fix the following plot?

```{r}
ggplot(daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(color = "blue")
```

# Splitting over facets

## Data prep

```{r}
daily <- daily %>%
  mutate(
    day = if_else(day_of_week %in% c("Sat", "Sun"), "Weekend", "Weekday"),
    temp = if_else(month %in% c("Jan", "Feb", "Mar", "Apr", "May", "Jun"),
                   "Cooler", "Warmer")
    ) %>%
  select(day, temp, everything())

daily
```

## facet_wrap

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides)) +
  geom_line() +
  facet_wrap( ~ day)
```

## facet_grid

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides)) +
  geom_line() +
  facet_grid(temp ~ day)
```

## facet_grid

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides)) +
  geom_line() +
  facet_grid(day ~ temp)
```

## Durham voters

```{r}
durham_voters %>%
  select(race_code, gender_code, age)
```

## Data prep

```{r}
durham_voters %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(), n_rep = sum(party == "REP"))
```

## Data prep

```{r}
durham_voters_summary <- durham_voters %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_all_voters = n(), n_rep_voters = sum(party == "REP")) %>%
  filter(gender_code %in% c("F", "M") & 
         race_code %in% c("W", "B", "A") &
         age != "Age < 18 Or Invalid Birth Date")
```

## facet_grid

```{r}
ggplot(durham_voters_summary, aes(x = age, y = n_all_voters)) +
  geom_bar(stat = "identity") +
  facet_grid(race_code ~ gender_code)
```

## Free scales

```{r}
ggplot(durham_voters_summary, aes(x = age, y = n_all_voters)) +
  geom_bar(stat = "identity") +
  facet_grid(race_code ~ gender_code, scales = "free_y")
```

## Facets + layers

```{r, echo = FALSE}
durham_voters_summary %>% 
  tidyr::pivot_longer(cols = starts_with("n_"), 
                      names_to = "voter_type", values_to = "n", 
                      names_prefix = "n_") %>%
  mutate(age_cat = as.numeric(as.factor(age))) %>%
  ggplot(aes(x = age, y = n, color = voter_type)) + 
    geom_point() + 
    geom_line(aes(x = age_cat)) +
    facet_grid(race_code ~ gender_code, scales = "free_y") +
    expand_limits(y = 0)
```

## Facets + layers

Using new tidyr function: `pivot_longer()`

```{r, eval = FALSE}
durham_voters_summary %>% 
  tidyr::pivot_longer(cols = starts_with("n_"), 
                      names_to = "voter_type", values_to = "n", 
                      names_prefix = "n_") %>%
  mutate(age_cat = as.numeric(as.factor(age))) %>%
  ggplot(aes(x = age, y = n, color = voter_type)) + 
    geom_point() + 
    geom_line(aes(x = age_cat)) +
    facet_grid(race_code ~ gender_code, scales = "free_y") +
    expand_limits(y = 0)
```

# Scales and legends

## Scale transformation

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_reverse()
```

## Scale transformation

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_sqrt()
```

## Scale details

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_continuous(breaks = c(0, 200, 500))
```

# Themes and refinements

## Overall themes

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme_bw()
```

## Overall themes

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme_dark()
```

## Customizing theme elements

```{r}
ggplot(data = daily, aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

## Your turn!

**Exercise:** Fix the axis labels in the following plot so they are at a 45 degree angle.

```{r}
ggplot(durham_voters_summary, aes(x = age, y = n_all_voters)) +
  geom_bar(stat = "identity") +
  facet_grid(race_code ~ gender_code, scales = "free_y")
```
